---
title: "EDA"
format: html
editor: visual
---
## Libraries

Before getting started lets read in the necessary libraries

```{r}
library("tidyverse")
```
## Task 1: Read in & Modify

Lets read in our data. The below code chunked was pulled from the `student-merge.R` file in our `files` folder and modified to reflect the folder structure of this project.

```{r}
d1=read.table("files/student-mat.csv",sep=";",header=TRUE)
d2=read.table("files/student-por.csv",sep=";",header=TRUE)

d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
d3
```

For the purpose of this homework, we want to read the data in the tidyverse way. We will use the joins provided the the source team.

One thing we notice is that there are other columns outside of those in the specified join that seem to match, which could result in duplication and highly dimensional data. Some examples include *guardian, freetime, and famrel*. In addition, some variables that appear to be cateogrical and should be cast as factors are cast as **dbl** type.

```{r}
math_data <- read_delim(file="files/student-mat.csv",delim=";",show_col_types = FALSE)
port_data <- read_delim(file="files/student-por.csv",delim=";",show_col_types = FALSE)
merged_data <- inner_join(math_data,port_data,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))

merged_data
```

Lets instead merge our data on all columns besides `G1`, `G2`, `G3`, `paid`, and `absences`. We know in our `colnames` output those NOT joined will have *.x* or *.y* following them, so we can confirm the appropriate columns were not included.

We also see that our tibble maintains the same # of observations

```{r}
joinCols <- colnames(math_data)[!(colnames(math_data) %in% c('G1','G2','G3','paid','absences'))]
merged_data <- inner_join(math_data,port_data,by=joinCols)
colnames(merged_data)
merged_data
```

Now to pick 4 variables that we can cast as factors across our 4 data frames. I think that address, studytime, famsup and activities will be interesting to look at.

To avoid repetitive code, we will write a function that takes in a vector of column names and applies to the tibbles. We'll apply that function across tibbles

```{r}
TibbleMultiFactor <- function (tibble,variables=NULL){
  returnTibble <- tibble %>%
      mutate(across(all_of(variables), as.factor))
  return(returnTibble)
}

variablesToFactor <- c('address','studytime','famsup','activities')

math_data <- TibbleMultiFactor(math_data,variables=variablesToFactor)
port_data <- TibbleMultiFactor(port_data,variables=variablesToFactor)
merged_data <- TibbleMultiFactor(merged_data,variables=variablesToFactor)
```

